DOpElib:\\ Differential Equations and Optimization Environment

### 2021-09-17 Examples/OPT/StatPDE/Example1/content.tex

In abstract form, we are given the following optimization problem:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Lagrangian:
\begin{equation*}
{\cal L}(q,u,z):=J(q,u) - a(q,u)(z)
\end{equation*}
Optimality system (KKT system):
\begin{align*}
a_u'(q,u)(\phi , z) &= J'_u(q,u)(\phi) \quad \forall \phi\in V \\
a_q'(q,u)(\chi , z) &= J'_q(q,u)(\chi) \quad \forall \chi\in Q \\
a(q,u)(\psi) &= 0 \qquad \forall\psi\in V
\end{align*}
or equivalently, in terms of the Lagrangian
\begin{align*}
{\cal L}'_u(q,u,z)(\phi) &= 0 \quad\forall \phi\in V \quad\text{(Adjoint Equation)} \\
{\cal L}'_q(q,u,z)(\chi) &= 0 \quad\forall \chi\in V \quad\text{(Gradient Equation)} \\
{\cal L}'_z(q,u,z)(\psi) &= 0 \quad\forall \psi\in V \quad\text{(State Equation)}
\end{align*}
The continuous problem is discretized by a standard Galerkin method using 
finite dimensional subspaces $Q_h \times V_h \subset Q\times V$:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Discrete saddle-point problems
\begin{align*}
a_u'(q_h,u_h)(\phi_h , z_h) &= J'_u(q_h,u_h)(\phi_h) \quad \forall \phi_h\in V_h \\
a_q'(q_h,u_h)(\chi_h , z_h) &= J'_q(q_h,u_h)(\chi_h) \quad \forall \chi_h\in Q_h \\
a(q_h,u_h)(\psi_h) &= 0 \qquad \forall\psi_h\in V_h
\end{align*}
\subsubsection{Solution process}
In this section, we briefly discuss the solution process for the 
optimization problem. For further details, we refer to the standard literature.
The unconstrained optimal control problem is reformulated as follows. 
We introduce the solution operator $S:Q\rightarrow V$ of the state equation.
Then:
\begin{equation*}
j(q):= J(q,S(q))\, \rightarrow \, \min , \quad a(q,S(q))(\Psi) = 0 \quad 
\forall \Psi \in V.
\end{equation*}
The local existence and sufficient regularity of $S$ is assumed. The 
necessary optimality conditions of first and second order are
\begin{equation*}
j'(q)(\delta q) = 0 , \quad j''(q)(\delta q,\delta q) \geq 0 \quad 
\forall\delta q\in Q.
\end{equation*} 
The derivatives of the reduced functional can be computed using the 
Lagrangian
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}
as already introduced. Let $q\in Q$, and the corresponding state $u=S(q)\in V$ 
be given. To calculate the derivative of the reduced cost functional $j$, 
we introduce the dual variable
$z\in V$ solving the\\[3mm]
%
\textbf{Dual equation}\\
\begin{equation*}
{\cal L}_u'(q,u,z)(\psi) = 0 \quad\forall\psi\in V.
\end{equation*}
Then,
\begin{equation*}
j'(q)(\delta q) = {\cal L}_q' (q,u,z)(\delta q) \quad \text{for }\delta q\in Q.
\end{equation*}
\\[2mm]
%
To calculate the second derivatives, we need to solve additional equations.\\
%
Let $\delta q\in Q$ be a given direction. Then we search $\delta u\in V$  
solving the\\[2mm]
%
\textbf{Tangent equation}\\
%
\begin{equation*}
{\cal L}_{qz}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uz}'' (q,u,z)(\delta
u,\phi) = 0
\quad\forall \phi\in V.
\end{equation*}
Further, we have an auxilliary\\[2mm]  
%
\textbf{Dual for Hessian equation}
to find $\delta z\in V$ solving
\begin{equation*}
{\cal L}_{qu}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uu}'' 
(q,u,z)(\delta u,\phi) + {\cal L}_{zu}'' (q,u,z)(\delta z,\phi) =0 
\quad\forall \phi\in V.
\end{equation*}

@@

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Simillarly, we can define the Hessian operator 
$H(q) \in \mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}

@@

\subsubsection{\texttt{main.cc}}
Finally, the main file of the optimization examples does not look very much 
different than for pure PDE computations - which is one of the crucial aims 
of our library. 
Here, instead of using a \texttt{pdeproblemcontainer}, we use 
now an \texttt{optproblemcontainer} which can assemble all additionally needed informations,
such as adjoint and tangent PDEs. 
Furthermore, we define \texttt{ReducedNewtonAlgorithm} and 
\texttt{ReducedTrustregion\underline{ }NewtonAlgorithm}
to solve the optimization problem with a linesearch and a trust-region Newton algorithm.
Of course one would be sufficient, but we wanted to show how to change optimization 
solvers easily using \texttt{DOpElib}.\\
Next, in the body of the main file, we introduce a second FE function for the 
control variable. Then, we define 
a \texttt{COSTFUNCTIONAL}. Finally, the problem is either solved 
by calling \texttt{Alg.Solve(q)} and/or the user might check if the 
derivatives are implemented correctly by calling 
\texttt{Alg.CheckGrads} or \texttt{Alg.CheckHessian}. The latter 
two functionalities are highly recommended to check your
implementation before wondering about your results.\\
Finally, this example uses a DWR-error estimator to estimate the 
error made in the cost functional. In contrast to the 
error estimation for PDEs here, we have to include the error in the
control by using the \texttt{HigherOrderDWRContainerControl}.

>>>>

In abstract form, we are given the following optimization problem:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Lagrangian:
\begin{equation*}
{\cal L}(q,u,z):=J(q,u) - a(q,u)(z)
\end{equation*}
Optimality system (KKT system):
\begin{align*}
a_u'(q,u)(\phi , z) &= J'_u(q,u)(\phi) \quad \forall \phi\in V \\
a_q'(q,u)(\chi , z) &= J'_q(q,u)(\chi) \quad \forall \chi\in Q \\
a(q,u)(\psi) &= 0 \qquad \forall\psi\in V
\end{align*}
or equivalently, in terms of the Lagrangian
\begin{align*}
{\cal L}'_u(q,u,z)(\phi) &= 0 \quad\forall \phi\in V \quad\text{(Adjoint Equation)} \\
{\cal L}'_q(q,u,z)(\chi) &= 0 \quad\forall \chi\in V \quad\text{(Gradient Equation)} \\
{\cal L}'_z(q,u,z)(\psi) &= 0 \quad\forall \psi\in V \quad\text{(State Equation)}
\end{align*}
The continuous problem is discretized by a standard Galerkin method using 
finite dimensional subspaces $Q_h \times V_h \subset Q\times V$:
\begin{equation*}
J(q,u) \, \rightarrow \, \min , \quad a(q,u)(\psi) = 0 \quad \forall\psi\in V
\end{equation*}
Discrete saddle-point problems
\begin{align*}
a_u'(q_h,u_h)(\phi_h , z_h) &= J'_u(q_h,u_h)(\phi_h) \quad \forall \phi_h\in V_h \\
a_q'(q_h,u_h)(\chi_h , z_h) &= J'_q(q_h,u_h)(\chi_h) \quad \forall \chi_h\in Q_h \\
a(q_h,u_h)(\psi_h) &= 0 \qquad \forall\psi_h\in V_h
\end{align*}
\subsubsection{Solution process}
In this section, we briefly discuss the solution process for the 
optimization problem. For further details, we refer to the standard literature.
The unconstrained optimal control problem is reformulated as follows. 
We introduce the solution operator $S:Q\rightarrow V$ of the state equation.
Then:
\begin{equation*}
j(q):= J(q,S(q))\, \rightarrow \, \min , \quad a(q,S(q))(\Psi) = 0 \quad 
\forall \Psi \in V.
\end{equation*}
The local existence and sufficient regularity of $S$ is assumed. The 
necessary optimality conditions of first and second order are
\begin{equation*}
j'(q)(\delta q) = 0 , \quad j''(q)(\delta q,\delta q) \geq 0 \quad 
\forall\delta q\in Q.
\end{equation*} 
The derivatives of the reduced functional can be computed using the 
Lagrangian
\begin{equation*}
{\cal L}(q,u,z) =J(q,u) - a(q,u)(z)
\end{equation*}
as already introduced. Let $q\in Q$, and the corresponding state $u=S(q)\in V$ 
be given. To calculate the derivative of the reduced cost functional $j$, 
we introduce the dual variable
$z\in V$ solving the\\[3mm]
%
\textbf{Dual equation}\\
\begin{equation*}
{\cal L}_u'(q,u,z)(\psi) = 0 \quad\forall\psi\in V.
\end{equation*}
Then,
\begin{equation*}
j'(q)(\delta q) = {\cal L}_q' (q,u,z)(\delta q) \quad \text{for }\delta q\in Q.
\end{equation*}
\\[2mm]
%
To calculate the second derivatives, we need to solve additional equations.\\
%
Let $\delta q\in Q$ be a given direction. Then we search $\delta u\in V$  
solving the\\[2mm]
%
\textbf{Tangent equation}\\
%
\begin{equation*}
{\cal L}_{qz}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uz}'' (q,u,z)(\delta
u,\phi) = 0
\quad\forall \phi\in V.
\end{equation*}
Further, we have an auxiliary\\[2mm]  
%
\textbf{Dual for Hessian equation}
to find $\delta z\in V$ solving
\begin{equation*}
{\cal L}_{qu}'' (q,u,z)(\delta q,\phi) + {\cal L}_{uu}'' 
(q,u,z)(\delta u,\phi) + {\cal L}_{zu}'' (q,u,z)(\delta z,\phi) =0 
\quad\forall \phi\in V.
\end{equation*}

@@

Moreover, we would like to work in the Hilbert space $Q$. However, the derivative 
$j'(q) \in H^*$ only. Hence, we need to calculate the Riesz representation 
for the gradient $\nabla j(q) \in H$ using the definition:
\[
 (\nabla j(q),\delta q)_Q = j'(q)(\delta q) \quad \forall\,\delta q\in Q.
\]
In the given example, $Q = L^2(\Omega)$ and hence the scalar product will be the
standard $L^2$-inner product. Similarly, we can define the Hessian operator 
$H(q) \in \mathcal L(Q,Q)$ by defining
\[
 (H(q)\tau q,\delta q)_Q = j''(q)(\tau q,\delta q) \quad \forall\,\delta q,\tau q\in Q.
\]
}

@@

\subsubsection{\texttt{main.cc}}
Finally, the main file of the optimization examples does not look very much 
different than for pure PDE computations - which is one of the crucial aims 
of our library. 
Here, instead of using a \texttt{pdeproblemcontainer}, we use 
now an \texttt{optproblemcontainer} which can assemble all additionally needed information,
such as adjoint and tangent PDEs. 
Furthermore, we define \texttt{ReducedNewtonAlgorithm} and 
\texttt{ReducedTrustregion\underline{ }NewtonAlgorithm}
to solve the optimization problem with a linesearch and a trust-region Newton algorithm.
Of course one would be sufficient, but we wanted to show how to change optimization 
solvers easily using \texttt{DOpElib}.\\
Next, in the body of the main file, we introduce a second FE function for the 
control variable. Then, we define 
a \texttt{COSTFUNCTIONAL}. Finally, the problem is either solved 
by calling \texttt{Alg.Solve(q)} and/or the user might check if the 
derivatives are implemented correctly by calling 
\texttt{Alg.CheckGrads} or \texttt{Alg.CheckHessian}. The latter 
two functionalities are highly recommended to check your
implementation before wondering about your results.\\
Finally, this example uses a DWR-error estimator to estimate the 
error made in the cost functional. In contrast to the 
error estimation for PDEs here, we have to include the error in the
control by using the \texttt{HigherOrderDWRContainerControl}.

### 2021-09-17 Examples/PDE/InstatPDE/Example10/content.tex

\subsubsection{General problem description}
We consider phase-field fracture propagation on a slit domain, with time dependent 
Dirichlet data and homogeneous Neumann data.
The main novelties are:
\begin{itemize}
\item Imposing a variational inequality constraint ($\frac{\partial \varphi}{\partial t} \leq 0$)
via a Langrage multiplier similar to  \cite{mang2020phase};
%\item Implementation of an augmented Lagrangian loop (not yet)
\item Stress splitting into tension and compression \`a
la \cite{MieWelHof10a}, see as well \cite{AmorMarigoMaurini2009}, located 
in 
\begin{verbatim}
stress_splitting.h
\end{verbatim}
\item Implementation of a Newton solver that 
allows for an increase of the residual for which, however,
no convergence results are available:
\begin{verbatim}
instat_step_modified_newtonsolver.h
\end{verbatim}
\item The localpde* files:
\begin{verbatim}
localpde_quasi_monolithic.h
\end{verbatim}
which provides the implementation of a quasi-monolithic scheme
(similar to\cite{HeWheWi15}).
\end{itemize}

The problem formulation as well as the handling of the irreversibility constraint is exactly the same as in Example 5.2.8.

The test case is the so-called single edged notched 
tension test. For further details about this problem it is referred
to \cite{MieHofWel2010,BoVeScoHuLa12}, and \cite{WeWiWo2014}.

% \subsubsection{Future improvements}
% Possible future improvements are:
% \begin{itemize}
% \item Implement the phase-field extrapolation in the displacement equation proposed in \cite{HeWheWi15};
% \item Implement the augmented Lagrangian loop proposed by
% Wheeler/Wick/Wollner, CMAME 2014 \cite{WeWiWo2014}. 
% \end{itemize}

% Need to implement extrapolation with two 
% previous time steps
% To be changed: 
% instatpdeproblem.h
% instat_step_newton_solver.h
%
% augmented Lagrangian solver loop
% -> currently simple penalization
% has to be implemented in instat_step_newton_solver.h
%
% Dokumentation in content.tex

>>>>

\subsubsection{General problem description}
In this example, we provide an implementation if the
The test case is the so-called single edged notched 
tension test. For further details about this problem it is referred
to \cite{MieHofWel2010,BoVeScoHuLa12}, and \cite{WeWiWo2014}.
For the choice of the boundary conditions see also~\cite[Section~5]{MangWallothWickWollner:2019}.

The formulation of the irreversibility constraint  is exactly the same
as in Example~\ref{PDE_phase_field_fracture}, however, the fracture is
not handled fully implicit. Instead the elasticity equation
couples with the phase-field from the previous time step, i.e., in
contrast to Example~\ref{PDE_phase_field_fracture} the following
system 
\begin{align*}
 \begin{aligned}
   \left(g(\varphi_h^{n-1})\left[2\mu E^+(u_h) + \lambda (\operatorname{trace}(E(u)))_+ \textbf{I}\right],\nabla w_h \right)\\
  + \left(2\mu \left(E(u_h)-E^+(u_h)\right),E(w_h)\right)\\
 + \left(\lambda (\operatorname{trace}(E(u)))_- \textbf{I},E(w_h)\right) =&\ 0,\\
 %\forall w_h\in&\ \mathcal{V}_h,\\
 %\left(g(\varphi_h) \nabla \cdot u_h,q_h\right) - \frac{1}{\lambda} \left(g(\varphi_h) p_h,q_h\right)=&\ 0,\\
 %\forall q_h\in&\ \mathcal{U}_h,\\
  (1- \kappa)\left(\varphi_h 2\mu E^+(u_h) + \lambda (\operatorname{trace}(E(u)))_+ \textbf{I} : E(u_h),\psi_h\right)\\
  + G_c \left(-\frac{1}{\epsilon} (1-\varphi_h),\psi_h\right)\\ 
  + G_c \epsilon \left(\nabla \varphi_h,\nabla\psi_h \right) - \left(\tau_h,\psi_h\right) =&\ 0,\\
  %\forall \psi_h \in&\ \mathcal{W}_h,\\
%\tau_h\geq&\ 0,\\
%\varphi_h^n - \varphi_h^{n-1} \leq&\ 0,\\
%\left(\tau_h,\varphi_h^n -\varphi_h^{n-1}\right)=&\ 0,
\left(\tau_h - \chi_h,\varphi_h -\varphi_h^{n-1}\right)\geq&\ 0,
\end{aligned}
\end{align*}
is solved using a complementarity formulation for the variational inequality.

### 2021-09-17 Examples/PDE/InstatPDE/Example9/content.tex

To use the Rothe-Discretization, we use a different DoF-Handler, i.e.,
we need to include
\begin{verbatim}
#include <basic/rothe_statespacetimehandler.h>
\end{verbatim}
To initialize the different meshes, we prepare a vector
\texttt{Rothe\_time\_to\_dof} which stores for each time-point a number
indicating the numer of the DoF-Handler to be used. In this example,
we want a different DoF-Handler at each time-point, hence we
initialize
\begin{verbatim}
std::vector<unsigned int> Rothe_time_to_dof(n_time_steps+1,0);

@@

The only other change to other programms is that now the refinement
needs to be given error indicators for each time-point, i.e., a
\verb|std::vector<dealii::Vector<float> >| where the outer
\texttt{std::vector} needs to have the size \texttt{n\_time\_steps+1}.
This is immeadiately given by the
\texttt{ResidualErrorContainer}. Hence, for refinement of the
different spatial meshes, we just have to call
\begin{verbatim}
const std::vector<dealii::Vector<float> > 
               error_ind(h1resc.GetErrorIndicators());
DOFH.RefineSpace(SpaceTimeRefineOptimized(error_ind));
\end{verbatim}

>>>>

To use the Rothe-Discretization, we use a different DoF-Handler, i.e.,
we need to include
\begin{verbatim}
#include <basic/rothe_statespacetimehandler.h>
\end{verbatim}
To initialize the different meshes, we prepare a vector
\texttt{Rothe\_time\_to\_dof} which stores for each time-point a number
indicating the number of the DoF-Handler to be used. In this example,
we want a different DoF-Handler at each time-point, hence we
initialize
\begin{verbatim}
std::vector<unsigned int> Rothe_time_to_dof(n_time_steps+1,0);

@@

The only other change to other programs is that now the refinement
needs to be given error indicators for each time-point, i.e., a
\verb|std::vector<dealii::Vector<float> >| where the outer
\texttt{std::vector} needs to have the size \texttt{n\_time\_steps+1}.
This is immediately given by the
\texttt{ResidualErrorContainer}. Hence, for refinement of the
different spatial meshes, we just have to call
\begin{verbatim}
const std::vector<dealii::Vector<float> > 
               error_ind(h1resc.GetErrorIndicators());
DOFH.RefineSpace(SpaceTimeRefineOptimized(error_ind));
\end{verbatim}

### 2021-09-17 Examples/description.tex
updated documentation
Renamed Example4_TimeSpace to Example8 and started Documentation
Added Example for the Obstacle problem
Added Example17 - Use of FEInterpolatedValues
Removed tmp files, added Copyright, added FIXMEs, updated description of Example 17

\documentclass[a4paper,cleardoubleempty]{scrreprt}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb, amsthm, commath}
\usepackage[pdftex,bookmarks=false,pdfpagelabels=true,plainpages=false]{hyperref}
\usepackage{booktabs}
\usepackage{makeidx}
\usepackage{url}
%\usepackage{sfmath}
\usepackage{color}
\usepackage{graphicx} 
\usepackage{enumerate}
\usepackage{listings}

@@

@@

@@

\subsection{Fully monolithic phase-field fracture propagation}

@@

>>>>

\documentclass[a4paper,cleardoubleempty]{scrreprt}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amssymb, amsthm, commath, bm}
\usepackage[pdftex,bookmarks=false,pdfpagelabels=true,plainpages=false]{hyperref}
\usepackage{booktabs}
\usepackage{makeidx}
\usepackage{url}
%\usepackage{sfmath}
\usepackage{color}
\usepackage{graphicx} 
\usepackage{enumerate}
\usepackage{listings}
\usepackage{tikz}

@@

\DeclareMathOperator{\tr}{tr} % new
\DeclareMathOperator{\Div}{div} % new
\newcommand{\bu}{\mathbf{u}}
\newcommand{\bv}{\mathbf{v}}

@@

\subsection{Parallel Solve for Laplace}
\label{PDE_parallel}
\input{PDE/StatPDE/Example15/content.tex}
\subsection{Obstacle Problem}
\label{PDE_obstacle}
\input{PDE/StatPDE/Example16/content.tex}
\subsection{Gradient Robust Discretization of Nearly Incompressible Elasticity}
\label{PDE_pressurerobust}
\input{PDE/StatPDE/Example17/content.tex}

@@

\subsection{Fully monolithic phase-field fracture propagation: Single Edged Notched Shear Test}

@@

\subsection{Heat Equation in 2D with Rothe-discretization}
\label{PDE_Instat_Heat_Rothe}
\input{PDE/InstatPDE/Example9/content.tex}
\subsection{Time lagged phase-field fracture propagation: Single Edged Notched Tension Test}
\label{PDE_SENT}
\input{PDE/InstatPDE/Example10/content.tex}
\subsection{Pressure-driven cavity - Sneddon's benchmark}
\label{PDE_Sneddon}
\input{PDE/InstatPDE/Example11/content.tex}
\subsection{Sneddon's benchmark layered for incompressible solids}
\label{PDE_Sneddon-incompressible}
\input{PDE/InstatPDE/Example12/content.tex}
